### 2021-11-29 chapter2/chapter.tex
first steps of numerics chapter

%!TEX root = ../thesis.tex
\chapter{A stochastic modelling approach to approximating fluid queues\label{sec: construction and modelling}}
In this chapter we develop a new discretisation of a fluid queue using a \emph{quasi-birth-and-death process with rational arrival process components} (QBD-RAP). In the following chapter, Chapter \ref{sec: conv} we prove the scheme converges weakly. To improve upon the discretisation of \cite{bo2013}, we argue that the shift from QBD to QBD-RAP is necessary. The QBD construction of \cite{bo2013} uses Erlang distributions to approximate the deterministic time that the fluid queue spends in an interval on the event that the phase is constant -- it is well known that the Erlang distribution is the distribution with the smallest variance in the class of Phase-type distributions of a given order \cite{as1987}, and in this sense is the best approximation to the determinstic behaviour we want to capture.

By choosing a matrix-exponential distribution with sufficiently small variance, the QBD-RAP approximation can be made arbitrarily accurate. Due to its stochastic interpretation, the QBD-RAP discretisation ensures solutions are positive, even when discontinuities are present. Unlike slope limiters and post-hoc filtering, for the QBD-RAP the positivity preserving nature is a property of the discretised operator.

The structure of this chapter is as follows. Section~\ref{sec: inspiration} motivates the idea using the QBD approximation of \cite{bo2014}. Sections \ref{sec: modelling} and \ref{sec: zero states} describe the modelling of certain events of the fluid queue with matrix exponential distributions to ultimately construct the behaviour of the QBD-RAP on the event that the QBD-RAP remains in a given level. Section~\ref{sec: dynamics} describes the dynamics of the construction and also constructs the level process of the QBD-RAP. Section~\ref{sec: boundary conditions} deals with modelling boundary behaviour. Section~\ref{sec: initial conditions} describes how to model initial conditions of the fluid queue. Section~\ref{sec: closing} introduces the concept of a \emph{closing operator} which maps the state space of the QBD-RAP to estimates of the density of the fluid queue.

\section{Inspiration and motivation}\label{sec: inspiration}
The inspiration for our approach stems from the QBD-based discretisation of \cite{bo2013}. A QBD can be seen as a two dimensional CTMC, \(\{(L(t),\phi(t))\}_{t\geq 0}\), where \(\{L(t)\}\) is the discrete level, and \(\{\phi(t)\}\) is the phase. \cite{bo2013} discretise the state space of the fluid into small intervals of width \(\Delta\); denote these by \(\mathcal D_k=[k\Delta,(k+1)\Delta]\). Their approximation captures the dynamics of the phase process exactly, but discretises the level process of the fluid queue. Their approximation supposes that when \(L(t)=k,\,\phi(t)=i\), then \(X(t)\approx k\Delta,\,\phi(t)=i\).

When in level \(k\) and phase \(i\), the approximation sees events at rate \(|T_{ii}| + |c_i|\Delta\). Upon an event, with probability \(T_{ij}/(|T_{ii}| + |c_i|\Delta)\) a change of phase from \(i\) to \(j\) is observed and the approximation remains in level \(k\); with probability \(|c_i|\Delta/(|T_{ii}| + |c_i|\Delta)\) a change of level occurs, to \(k+1\) if \(c_i>0\) and \(k-1\) if \(c_i<0\). The generator of their QBD approximation (for an unbounded fluid queue) is 
\[\bs B = \left[ \begin{array}{cccc} \ddots & \ddots & \ddots & \\ \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) & \\  & \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) \\ & \ddots & \ddots & \ddots \end{array} \right]\]
\begin{align}
\bs B_0 &= \bs T- \bs C\Delta, && \bs B_{-1} = \left[\begin{array}{ccc} \bs 0 & & \\ & \bs C_-\Delta & \\ & & \bs 0 \end{array}\right], && \bs B_{+1} = \left[\begin{array}{ccc} \bs C_+\Delta & & \\ & \bs 0 & \\ & & \bs 0\end{array}\right].
\end{align}
Bean and O'Reilly then take \(\Delta \to 0\) and show that the approximation converges weakly to the fluid queue. It seems that this is the best we can do if we want to keep the interpretation of the approximation as a QBD. We now elaborate on this point.

For a given \(\Delta\), the discretisation of Bean and O'Reilly supposes that, when the phase is \(i\) and on the event of no change of phase, the sojourn time in a given level has an exponential distribution with rate \(|c_i|\Delta\). However, the corresponding sojourn time of the fluid queue is deterministic: given \(X(t) = x \in\calD_k\), then \(\{X(t)\}\) will leave \(\calD_k\) in exactly \(((k+1)\Delta - x)/|c_i|\) units of time if \(c_i>0\), and \((x-k\Delta)/|c_i|\) units of time if \(c_i<0\), on the event that the phase does not change before this time. We can extend the QBD model of Bean and O'Reilly to model this determinism more accurately by supposing that the sojourn times in each interval have Erlang distributions (rather than exponential). However, we effectively realise the same QBD approximation as the original, but on a finer discretisation. For example, using Erlang distributions of order 2 we would get a generator of the QBD approximation
\[\bs B = \left[ \begin{array}{cccc} \ddots & \ddots & \ddots & \\ \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) & \\  & \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) \\ & \ddots & \ddots & \ddots \end{array} \right]\]
\begin{align}
\bs B_0 &= \bs T\otimes I_2 - \left[ \begin{array}{ccc} \left[\begin{array}{cc} -\Delta/2 & \Delta/2 \\ 0 & -\Delta/2 \end{array}\right] \otimes \bs C_+ & & \\ & \left[\begin{array}{cc} -\Delta/2 & 0 \\ \Delta/2 & -\Delta/2 \end{array}\right] \otimes C_- & \\ & & \bs 0 \end{array}\right], 
\end{align}
\begin{align}
\bs B_{-1} &= \left[\begin{array}{ccc} \bs 0 & & \\ & \bs C_-\otimes \left[\begin{array}{cc} 0 & \Delta/2 \\ 0 & 0 \end{array}\right] & \\ & & \bs 0 \end{array}\right],
&&\bs B_{+1} = \left[\begin{array}{ccc} \bs C_+\otimes \left[\begin{array}{cc} 0 & 0 \\ \Delta/2 & 0 \end{array}\right] & & \\ & \bs 0 & \\ & & \bs 0\end{array}\right].
\end{align}
But this is the same discretisation we would get if we were to use intervals of width \(\Delta/2\) in the QBD approximation of \cite{bo2013}, with the rows/columns reordered. Indeed, this appears to be the best we can do with a QBD approximation as the Erlang distribution is known to be the least variable Phase-type distribution for a given order \citep{as1987} and therefore gives the closest approximation to the deterministic behaviour we are trying to approximate. This necessitates the extension to QBD-RAPs, whereby we can find more concentrated distributions than any Phase-type and retain enough stochastic interpretation and matrix-analytic tools.

Here, the construction of the approximating QBD-RAP is developed intuitively, from a stochastic modelling perspective. The key observation is that, given \(\{\varphi(t)\}\) is constant, then \(\{X(t)\}\) moves deterministically. Upon discretising the state space of the level process into intervals of width \(\Delta\), then, given \(X(t)\), the distribution of time it takes for \(\{X(t)\}\) to leave a given interval on the event that \(\{\varphi(t)\}\) remains in the same phase, is deterministic. We model this deterministic behaviour approximately by concentrated matrix exponential distributions (matrix exponential distributions with low-variance) \citep{ert2006,hstz2016,ert2006,hht2020,mt2021}. We construct a level process for the QBD-RAP and correspond it to a discretisation the level process of the fluid queue. This is similar to \cite{bo2013} where the level process of their QBD corresponds a discretisation of the fluid level. Unlike \cite{bo2013}, however, here we do not take \(\Delta \to 0\). Rather, we suppose that the variance of the concentrated matrix exponential distributions we use to approximate deterministic events gets small. % This necessarily means that the order of the matrix exponential increases.

It turns out that the corresponding orbit process \(\{\bs A(t)\}\) can be seen to ``track", approximately, how far \(X(t)\) is from the left of a given interval when the phase is in \(\mathcal S_+\), or how far from the right of the interval \(X(t)\) is when the phase is in \(\mathcal S_-\). This leaves us with two problems. 1) on a transition from \(\mathcal S_+\) to \(\mathcal S_-\) or \(\mathcal S_-\) to \(\mathcal S_+\) how must \(\bs A(t)\) jump to retain this information about where \(X(t)\) is within a given interval. 2) how to use the orbit position at time \(t\), \(\bs A(t)\), to obtain an approximation for the density of \(X(t)\). Answers to both problems can be derived from the \emph{residual time} of a matrix exponential distribution. Let us now formalise these concepts some more.

\section{Time to exit an interval}\label{sec: modelling}
Consider the partition of the state space of the level of a fluid queue, \([0,M]\) into \(K\) intervals of width \(\Delta = M/(K+1)\), specifically \(\mathcal D_{k,i} = [k\Delta,(k+1)\Delta)\) if \(i\in\mathcal S_+\) and \(\mathcal D_{k,i} = (k\Delta,(k+1)\Delta]\) if \(i\in\mathcal S_-\), \(k=0,\dots,K+1\). The distinction between \(\mathcal D_{k,i}\) for \(i\in\mathcal S_+\) and \(\calS_-\) is a technical one and will be discussed later. For now, one reason we might want to specify the discretisation in this way, is that it ensures the stochastic process 
\begin{align}
	\displaystyle \left\{\sum_{k\in\mathcal K}\sum_{i\in\calS} k1(X(t)\in\calD_{k,i})\right\}_{t\geq 0} \label{eqn: kfj}
\end{align}
is right-continuous with left limits (c\'adl\'ag). The expression (\ref{eqn: kfj}) is the discrete process of the fluid-queue which the level process of the QBD-RAP will approximate. Let \(y_k = k\Delta\), \(k=0,\dots,K+1\) and \(\calD_k = [y_k,y_{k+1}]\). We now look to model the behaviour of \(\{X(t)\}\) on an interval \(\calD_k\) over the time \([t,t+u]\) given information about the phase process over this time.

\subsection{Modelling the residual time to exit on no change of phase} 
We first consider no change of phase in \([t,t+u]\). Suppose that \(\varphi(t)=i\in\mathcal S_+\) and \(X(t) = x\in\mathcal D_{k,i}\). Observe that on no change of phase, \(\varphi(t+s) = i\), \(s\in[0,u]\), and at time \(t+s\) the level of the fluid queue is given by \(X(t+s) = x+c_i s\), \(s\in[0,u]\). Further, for \(u>(y_{k+1}-x)/c_i\) the level process \(\{X(t)\}\) leaves the interval \(\mathcal D_{k,i}\) at exactly time \(t+(y_{k+1} - x)/c_i\). Similarly, for \(i\in\mathcal S_-\), \(\{X(t)\}\) leaves the interval \(\mathcal D_{k,i}\) at exactly time \(t+(x-y_k)/|c_i|\) on the event that there is no change of phase on \([t,t+u]\) for \(u>(x-y_k)/|c_i|\).

To approximate the deterministic behaviour observed above, consider a matrix-exponential distribution \(Z\sim ME(\bs \alpha, \bs S, \bs s)\) with mean \(\Delta\) and low variance. As we shall see later, the assumption that \(Z\) has low variance will allow us to claim that \(Z\) can be used to approximate deterministic behaviour. Let \(Z_i\sim ME(\bs \alpha, \bs S_i, \bs s_i)\) where \(\bs S_i = |c_i|\bs S\), \(\bs s_i=-\bs S_i\bs e\), \(i\in\calS\). The random variable \(Z_i\) has mean \(\Delta/|c_i|\).

Suppose first that \(X(t)=y_k\) and \(\varphi(t)=i\in\mathcal S_+\), so that \(X(t)\) will leave \(\mathcal D_{k,i}\) in exactly \(\Delta/|c_i|\) units of time on the event that the phase process of the QBD-RAP remains in \(i\) for at least this amount of time. A sensible approximation for this deterministic event is to choose the value of the orbit process at time \(t\) to be \(\bs \alpha\), \(\bs A(t)=\bs\alpha\). With this choice, the distribution of time until the next event of the QBD-RAP, on the event that the phase process remains constant will be the distribution of \(Z_i\). Observe, 
\begin{align*}
	\mathbb P(Z_i \in ((\Delta-\varepsilon)/|c_i|, (\Delta+\varepsilon)/|c_i|) ) 
	&= \mathbb P(Z\in (\Delta-\varepsilon, \Delta + \varepsilon))
%	\\&= \cfrac{\mathbb P(Z_i-u\in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon))}{\mathbb P(Z_i-u > 0)} 
%	\\&\geq \mathbb P(Z-u\in (\Delta-u-\varepsilon, \Delta-u + \varepsilon))
	\\&\geq 1-\cfrac{\var(Z)}{\varepsilon^2},
\end{align*}
since, by Chebyshev's inequality, \(\mathbb P(Z\in (\Delta-\varepsilon, \Delta + \varepsilon)) \geq 1- \cfrac{\var(Z)}{\varepsilon^2}\). Choosing \(\varepsilon = \var(Z)^{1/3}\), then 
\[\mathbb P(R_i(u) \in ((\Delta-\varepsilon)/|c_i|-u, (\Delta+\varepsilon)/|c_i|) )\geq 1-\var(Z)^{1/3} \approx 1, \]
when \(\var(Z)\) is small. Hence, when the variance of \(Z\) (equivalently \(Z_i\)) is low, then with high probability the time until the next event of the QBD-RAP, given the phase process remains constant will occur at approximately \(\Delta/|c_i|\).

Now, let \(R_i(u)\) be the residual time, \(R_i(u) = (Z_i-u)1\{Z_i-u>0\}\), \(i\in\calS\). After \(u\in[0,\Delta/|c_i|)\) amount of time has elapsed, on the event that \(\varphi(t+s)=i\), for all \(s\in[0,u)\), then \(X(t+u) = y_k + c_i u\). If the phase remains \(i\) for a further \(\Delta/|c_i|-u\) amount of time, then \(\{X(t)\}\) will leave \(\calD_{k,i}\) at exactly this time. Also at time \(t+u\), given \(Z_i>u\), the distribution of the residual time, \(R_i(u)\), has density 
\[f_{R_i(u)}(r) = \cfrac{\bs\alpha e^{\bs S_i(u+ r) } \bs s_i}{\bs\alpha e^{\bs S_iu } \bs e} = \bs A(t+u) e^{\bs S_i r}  \bs s_i.\]
Here, the event \(Z_i>u\) approximates the event \(X(t+u)\in\calD_{k,i}\) and we want the residual time, \(R_i(u)\), to approximate the time until \(\{X(t)\}\) leaves \(\calD_{k,i}\). That is, we want \(R_i(u)\) to approximate a deterministic random variable at \(\Delta/|c_i| - u\). To this end, we observe that for \(\varepsilon>0\) and \(u<(\Delta-\varepsilon)/|c_i|\), then
\begin{align*}
	&\mathbb P(R_i(u) \in ((\Delta-\varepsilon)/|c_i|-u, (\Delta+\varepsilon)/|c_i|-u) ) 
	\\&= \mathbb P(Z_i\in ((\Delta-\varepsilon)/|c_i|, (\Delta+\varepsilon)/|c_i|) ) 
	\\&= \mathbb P(Z\in (\Delta-\varepsilon, \Delta + \varepsilon))
%	\\&= \cfrac{\mathbb P(Z_i-u\in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon))}{\mathbb P(Z_i-u > 0)} 
%	\\&\geq \mathbb P(Z-u\in (\Delta-u-\varepsilon, \Delta-u + \varepsilon))
	\\&\geq 1-\cfrac{\var(Z)}{\varepsilon^2}.
\end{align*}
Choosing \(\varepsilon = \var(Z)^{1/3}\), then 
\[\mathbb P(R_i(u) \in ((\Delta-\varepsilon)/|c_i|-u, (\Delta+\varepsilon)/|c_i|) )\geq 1-\var(Z)^{1/3} \approx 1, \]
when \(\var(Z)\) is small. 
That is, when the variance of \(Z\) (equivalently \(Z_i\)) is low, the residual time \(R_i(u)\) will be concentrated around \(\Delta/|c_i| - u\), as required. Figure \ref{fig:residual distributions} gives an example of a density function for a \(Z_i\) with mean \(\Delta=1\) and \(c_i=1\), as well as the density function of \(R_i(0.3)\) conditional on \(Z_i>0.3\) and \(R_i(0.6)\) conditional on \(Z_i>0.6\), for comparison. Observe that the density of the residual life \(R_i(0.3)\) conditional on \(Z_i>0.3\) is concentrated around \(\Delta-0.3 = 0.7\) and, similarly the density of the residual life \(R_i(0.6)\) conditional in \(Z_i>0.6\) is concentrated around \(\Delta-0.6 = 0.4\).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{chapter2/figs/ME_residual_life_density.pdf}
\caption{The density function for a \emph{concentrated matrix exponential of order 21} from \citep{hht2020} (blue) and corresponding density functions of the residual lives, \(R_{0.3}\) (red), \(R_{0.6}\) (blue). Observe how the density function of the \(Z_i\) (blue) to approximate a point mass at \(\Delta=1\), while the density functions of \(R_{0.3}\) (red) and \(R_{0.6}\) (blue) approximate point masses at \(0.7\) and \(0.4\), respectively. }
\label{fig:residual distributions}
\end{figure}

For notational convenience, define the row-vector-valued function \(\bs k(t):\mathbb R\to \mathcal A\), 
\begin{equation}
	\bs k(t) = \cfrac{\bs \alpha e^{\bs S t}}{\bs \alpha e^{\bs S t}\bs e}, \label{eqn: orbit fun def}
\end{equation}
and the operator \(\bs K(t): \mathcal A\to \mathbb R\), \(t\geq 0\), 
\begin{equation}
	\bs a \bs K(t) = \cfrac{\bs a e^{\bs S t}}{\bs a e^{\bs S t}\bs e}, \label{eqn: operator fun def}
\end{equation}
for any \(a\in\mathcal A\).

We interpret the position of the orbit \(\bs A(t+u) = \bs k(|c_i|u)\) as corresponding to the fact that \(X(t+u)\) is \(|c_i|u\) units from the left-hand boundary of the interval \(\calD_{k,i}\). This gives a heuristic argument as to how we can model the sojourn times in a given interval \(\mathcal D_{k,i}\) on the event that the phase does not change.

We can apply analogous arguments to heuristically develop a model for the sojourn time of the fluid queue in an interval, \(\calD_{k,i}\), \(i\in\calS_-\), on the event that the phase does not change, given the fluid is \(X(t)=y_{k+1}\). For \(i\in\calS_-\), the orbit position \(\bs A(t+u) = \bs k(|c_i|u)\) is interpreted as corresponding to \(X(t)\) being \(|c_i|u\) units from the right-hand boundary of the interval \(\calD_{k,i}\), \(y_{k+1}\).

\subsection{On a change of phase from \(i\in\calS_+\) to \(j\in\calS_+\)} 
%Let \(Z_j\sim ME(\bs\alpha,\bs S_j)\) be a matrix exponential random variable with mean \(\Delta/|c_j|\) and let \(R_j(u)\) be the residual time \(R_j(u) = Z_j-u \mid\{Z_j-u>0\}\). %At time \(t+u\) the value of the orbit is \(\bs A(t+u) =  \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\), which can be equivalently written as \(\bs A(t+u) =  \cfrac{\bs\alpha e^{\bs S_ju/|c_i| } }{\bs\alpha e^{\bs S_ju/|c_i| } \bs e}\).

Let \(E\) be the event that, at time \(t\) the orbit of the QBD-RAP is \(\bs A(t) = \bs \alpha\), the phase process is \(\varphi(t) = i\in\calS_+\), and there are no events of the QBD-RAP before time \(t+u\). On the event \(E\) the orbit position at time \(t+u\) will be 
\begin{align}
	\bs A(t+u) = \bs k(|c_i|u) = \cfrac{\bs \alpha e^{\bs S_i u}}{\bs \alpha e^{\bs S_i u}\bs e}.\label{eqn: orbit i}
\end{align}
Correspondingly, on the event that at time \(t\) the level process of the fluid queue is \(X(t)=y_{k}\), the phase is \(\varphi(t)=i\) and there are no change of phase by time \(t+u\), then \(X(t+u)=y_{k}+c_iu\). That is, \(X(t+u)\) is \(c_iu\) units from the left-hand edge of \(\calD_k\).

If, at time \(t+u\) there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_+\), then we need to map the orbit position \(\bs A(t+u^-) = \bs k(|c_i|u)\) to an orbit position which corresponds to the being \(c_iu\) units from the left-hand edge of \(\calD_k\) in phase \(j\). Call this mapping \(\bs D(i,j)(\cdot)\). As noted in Section \ref{sec: qbd-rap}, the mapping must be linear, 
\(\bs D(i,j)(\bs a) = \bs a \bs D(i,j),\, \bs a \in\mathcal A,\)
for some matrix \(\bs D(i,j)\). The matrix \(\bs D(i,j)\) is a modelling choice.

So that the process is a QBD-RAP, the matrix \(\bs D(i,j)\) must have the property that, for any \(\bs a\in\mathcal A\), \((\bs a\bs D(i,j), \bs S)\) is a representation of a matrix exponential distribution. We would also like the matrix \(\bs D(i,j)\) to have the property \(\bs D(i,j) \bs e = \bs e\) as this will mean that the rate at which a change of phase from \(i\) to \(j\) of the QBD-RAP occurs proportional to 
\[\bs A(t+u) \bs D(i,j)\bs e = 1,\]
which is constant and therefore the distribution of time until a change from phase \(i\) to \(j\) is exponential. We can use this to show, for certain models, the distribution of the phase process of the fluid queue and the distribution of the phase process of the QBD-RAP share the same distribution.

An orbit position which corresponds to the level process of the QBD-RAP beign \(c_iu\) units to the right of \(y_k\) in phase \(j\in\calS_+\) is 
\[\bs A(t+u) = \cfrac{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|)}}{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|) u}\bs e}.\] 
To see this, consider the event that at time \(t\) the orbit process of the QBD-RAP is \(\bs A(t) = \bs \alpha\), the phase process is \(\varphi(t) = j\in\calS_+\), and there are no events of the QBD-RAP before time \(t+|c_i|u/|c_j|\). On this event, the orbit position at time \(t+u\) will be 
\[\displaystyle \bs A(t+u) = \cfrac{\bs \alpha e^{\bs S_j (|c_i|u)/|c_j|}}{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|) u}\bs e}.\] 
Correspondingly, on the event that at time \(t\) the level process of the fluid queue is \(X(t)=y_{k}\), the phase is \(\varphi(t)=j\) and there are no change of phase by time \(t+|c_i|u/|c_j|\), then \(X(t+u)=y_{k}+c_j|c_i|u/|c_j| = y_k+c_iu\). That is, \(X(t+u)\) is \(c_iu\) units from the left-hand edge of \(\calD_k\).

Now, observe 
\[\cfrac{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|)}}{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|) u}\bs e} = \cfrac{\bs \alpha e^{\bs S_iu}}{\bs \alpha e^{\bs S_i u}\bs e} = \bs k(|c_i|u),\] 
which is exactly (\ref{eqn: orbit i}), the orbit position on the event \(E\). Hence, it is sensible to choose \(\bs D(i,j)=\bs I\).

% Consider the residual time \(R_j(|c_i|u/|c_j|)\), which is the distribution of time until \(Z_j-|c_i|u/|c_j|\) occurs, on the event \(Z_j>|c_i|u/|c_j|\). The initial vector associated with the residual time \(R_j(|c_i|u/|c_j|)\) is \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\). This is the initial vector which is obtained if at time \(t\) the phase is \(\varphi(t)=j\) and the orbit is \(\bs A(t)=\bs \alpha,\) so that the QBD-RAP has just entered level \(k\), and there is no change of phase by time \(|c_i|u/|c_j|\). Above, we argued that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) can be seen to correspond to the event that \(X(t+u)\) is \(|c_j|\left(|c_i|u/|c_j|\right) = |c_i|u\) units greater than \(y_k\). 
% %We argued earlier that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) is the position of the orbit on the event that at time \(t\) there is a change of level so that \(\bs A(t)=\bs \alpha\), and the phase is \(\varphi(t)=j\), followed by no change of phase by time \(t+|c_i|u/|c_j|\). The analogous event of the fluid queue is that at time \(t\) the phase is \(\varphi(t)=j\), the fluid level is \(X(t)=y_k\), so that the fluid level has just entered an interval \(\calD_k\) at time \(t\), and there is no change of phase by time \(t+|c_i|u/|c_j|\), in which case the fluid level will be \(X(t+|c_i|u/|c_j|)=y_\ell + |c_j|\left(|c_i|u/|c_j|\right)=y_\ell + |c_i|u\). Therefore, we claim that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) is the correct position of the orbit to approximate the fact that \(\{X(t)\}\) is \(|c_i|u\) units to the right of the point \(y_k\).

% Notice that we can re-write the vector 
% \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) as \(\cfrac{\bs \alpha e^{\bs S_iu } }{\bs \alpha e^{\bs S_iu} \bs e }\). This is the orbit position on the event that at time \(t\) there is a change of level, so that \(\bs A(t)=\bs \alpha\), and the phase is \(\varphi(t)=i\), followed by no change of phase by time \(t+u\). We have argued that this is the orbit position which approximates the fact that \(\{X(t)\}\) is \(|c_i|u\) units greater than \(y_k\).

% Therefore, at time \(t+u\), if there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_+\) then the orbit position should not change as in both cases the orbit position \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) corresponds to an approximation to \(X(t)\) being \(|c_i|u\) units greater than \(y_k\).

%Therefore, suppose that at time \(t\), \(X(t)=y_k\), and that at time \(t+u\), \(u\in[0,\Delta/|c_i|)\), before \(\{X(t)\}\) exits the band \(\calD_{k}\), there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_+\). At time \(t+u\) the position of the fluid level is \(X(t+u) = y_k+c_iu\), which is \(|c_i|u\) units to the right of \(y_k\) and \(\Delta-c_iu\) to the left of \(y_{k+1}\), the right-hand boundary of \(\calD_{k,i}\). The residual time until the level of the fluid queue leaves \(\calD_{k,j}\), on the event that it remains in phase \(j\) until this time, is \((\Delta - |c_i|u)/|c_j|\).

Moreover, the residual time \(R_j(|c_i|u/|c_j|)\) has density 
\[f_{R_j(|c_i|u/|c_j|)}(r) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}e^{\bs S_j r}\bs s_j,\]
%and corresponds to an orbit position of \(\bs A(t+u) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\). 
and 
\begin{align}
	&\mathbb P(R_j(|c_i|u/|c_j|) \in ((\Delta-|c_i|u-\varepsilon)/|c_j|, (\Delta-|c_i|u + \varepsilon)/|c_j| )) \nonumber
	\\&=\mathbb P(Z_j\in ((\Delta-\varepsilon)/|c_j|, (\Delta + \varepsilon)/|c_j|)) 
	\\&=\mathbb P(Z\in (\Delta-\varepsilon ,  \Delta + \varepsilon )) 
	% \\&=  \mathbb P(Z_j\in (\Delta/|c_j|-\varepsilon, \Delta/|c_j| + \varepsilon)) 
%	\\&= \cfrac{\mathbb P(Z_i-u\in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon))}{\mathbb P(Z_i-u > 0)} 
%	\\&\geq \mathbb P(Z-u\in (\Delta-u-\varepsilon, \Delta-u + \varepsilon))
%	\\&\geq \cfrac{1-\cfrac{\var(Z_j)}{\varepsilon^2}}{1} 
	\\&\geq 1-\var(Z)^{1/3} \approx 1.
\end{align}
Hence, when the variance of \(Z\) is low, the residual time, \(R_j(|c_i|u/|c_j|)\), is concentrated around \((\Delta - |c_i|u)/|c_j|\), as required.

% Therefore, in our model, upon the event of a change from \(i\in\calS_+\) to \(j\in\calS_+\) at time \(t+u\) the orbit should be \(\bs A(t+u) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\). That is, the initial condition for the next segment of the orbit process is just the current orbit position and the orbit should start to evolve according to the matrix \(\bs S_j\).

Analogous arguments suggest that the same applies for changes of phase from \(i\in\calS_-\) to \(j\in\calS_-\).

\subsection{On a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\)} 
Now suppose \(X(t)=y_k\), \(\varphi(t)=i\in\calS_+\), and the phase remains in state \(i\) until there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), \(u\in[0,\Delta/|c_i|)\). %We have reasoned, intuitively, that, for \(i\in\calS_+\) and for \(x=|c_i|u\in[0,\Delta)\), the orbit position \(\bs k(x)\) corresponds to the fluid level position \(X(t) = y_k+x\), that is \(x\) units greater than \(y_k\).  
As before, we need to find a matrix \(\bs D(i,j)\) to map the orbit position on the event \(E\) when there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), from \[\bs A(t+u^-) = \bs k(|c_i|u^-),\] 
to 
\[\bs A(t+u) = \cfrac{\bs k(u^-)\bs D(i,j)}{\bs k(u^-)\bs D(i,j)\bs e}=\cfrac{\bs\alpha e^{\bs S_iu} \bs D(i,j)}{\bs\alpha e^{\bs S_iu}\bs D(i,j)\bs e}\]
in such a way that the orbit position after the jump corresponds, in some way, to the fluid level being a distance of \(|c_i|u\) from \(y_k\) in phase \(j\).

Again, the matrix \(\bs D(i,j)\) is a modelling choice. We first discuss how we might choose the matrix \(\bs D(i,j)\) for when the matrix-exponential \(Z\) is a phase-type distribution.

\paragraph{The Phase-type case}
If \(Z\sim ME(\bs \alpha, \bs S)\) is chosen to be a phase-type distribution then \(Z\) has the interpretation as the distribution of time until absorption of a finite-state continuous-time Markov chain with transient states \(\{1,\dots,p\}\) and a single absrobing state. The sub-generator matrix describing the dynamics of the Markov chain on transient states is \(\bs S\), and \(\bs \alpha\) is an initial probability distribution over the transient states. Let \(\{J(t)\}\) be the Markov chain associated with the phase-type distribution.

In the discussions above, we have relied on the relationship between the event \(E\) and the orbit position \(\displaystyle \bs k(|c_i|u)\). The relationship allows us to associate this orbit position with the level of the fluid queue. For phase-type distributions the vector \(\bs k(|c_i|u)\) is the vector of posterior probabilities \(\left[\mathbb P(J(u)=k \mid E)\right]_{k\in\{1,...,p\}}\). We can use this vector of posterior probabilities in the same way as we do for matrix-exponential distributions and QBD-RAPs. However, the orbit interpretation forgoes the presence of the phase process \(\{J(t)\}\). Here, we will use the phase process \(\{J(t)\}\) to derive a choice of \(\bs D(i,j)\).

>>>>

%!TEX root = ../thesis.tex
\chapter{Approximating fluid queues with the discontinuous Galerkin method \label{ch:galerkin}} 
% PREFACE WITH CONTRIBUTIONS 
% TUTORIAL: HOW TO -- AS IN PEPER
% PROBLEMS AND SOLUTIONS 
% 	NON-NEGATIVITY AND OSCILLATIONS 
%	FILTERING AND LIMITING 
%	NOT A SOLUTION IN OUR CONTEXT & MOTIVATION FOR A NEW METHOD -- THE QBD-RAP
